\bf On Devon Allen's Disqualification at the 2022 World Track and Field Championships

### 2024-11-25 Manuscript/manuscript.tex
minor edits and comments in manuscript

World Athletics (formerly IAAF) uses certified SIS to measure RTs, yet
variation in technology persist. Discussions at online forums such as
\url{www.LetsRun.com} questioned consistencies in the SIS as a
contributing factor to RT anormalies \citep{johnson2022data,
  johnson2022was}. Historically, ``loud gun'' systems caused signal delays
for athletes in outer lanes  due to the speed of sound, an issue
addressed with the introduction of ``silent gun'' systems in 2010,
which electronically synchronize sound delivery to all athletes
\citep{tonnessen2013reaction}. Despite these advances, variability
persists due to differences in sensor technologies, such as force
transducers and accelerometers, and inconsistencies in event detection
algorithms \citep{willwacher2013novel}. For example, simple
force-threshold systems may delay RT detection by up to 26~ms compared
to more sophisticated methods \citep{pain2007sprint}. These findings
emphasize the need for standardized certification protocols to reduce
discrepancies and ensure fairness in RT measurements, as recently
reviewed by \citet{milloz2021sprint}.

The 0.1-second disqualification threshold, introduced in the 1990s,
has been the subject of significant debate. Partly based on limited
data from Finnish national-level athletes~\cite{mero1990reaction}, the
threshold may not adequately represent the capabilities of elite
sprinters. Controlled experiments have shown that reaction times below
0.1 seconds are physiologically plausible \citep{pain2007sprint,
  komi2009iaaf}, while retrospective analyses of competition data
often advocate for raising the threshold~\cite{brosnan2017effects,
  lipps2011implications}. Stricter false-start rules, introduced to
minimize race disruptions, have also discouraged sprinters from
attempting faster starts, which may artificially inflate RTs recorded
in competition~\cite{haugen2013effect}. This divergence between
experimental findings and competition-based analyses illustrates the
complexity of defining a universally fair threshold. Addressing these
debates requires modern data collection and advanced methodologies to
ensure equity and consistency in elite
competition~\cite{milloz2021sprint}.

This paper addresses two primary objectives from a statistical
perspective, using modern methodologies to analyze historical
data. First, we investigate whether reaction times (RTs) at the 2022
World Championships were significantly different from other
competitions, focusing on athletes who competed in multiple
events. Using a matched-pairs design, we compare RTs across the 2022,
2019, and 2023 World Championships, as well as the 2022 national-level
competitions. This approach isolates the effect of the competition
year while controlling for individual performance. The clustered
nature of the data, with the goal of assessing differences across
events within athletes, was analyzed using a rank-based comparison
approach for clustered data \citep{datta2005rank}. Second, we evaluate
the appropriateness of the 0.1-second disqualification threshold by
modeling RTs from World Championships held from 1999 onward. A
generalized Gamma (GG) distribution with random effects for both venue and
heat was applied within the framework of the generalized additive
model for location, scale, and shape (GAMLSS)
\citep{rigby2005generalized, stasinopoulos2024generalized}.
This model enables estimation of the
probability of RTs falling below the threshold, providing a
statistical assessment of RT consistency and the validity of the
current threshold.

The rest of this paper is organized as follows. Section~\ref{sec:Data}
describes the data collection and datasets used for the two
objectives. Section~\ref{sec:rank} outlines the rank-based comparison
methods for objective~1, rank-based comparison for clustered data with
subunit-level gruping to analyze differences in RTs across different
competitions. Section~\ref{sec:gamlss} details the methods for
objective~2, the GG model with random effects to evaluate the appropriatenes
of the 0.1-second threshold. Section~\ref{sec:Results} presents the
results of both objectives and discusses the implications.
Finally, Section~\ref{sec:concludingremarks} highlights the
paper’s impact, limitations, and potential for future research.
All data and code for our analysis are provided in the supplementary
materials.

@@

The final dataset consisted of RTs from athletes who competed
in both national and international competitions. RTs from
preliminary heats, semifinals, and finals were included to ensure
sufficient data for analysis. Excluding preliminary heats would have
significantly reduced the number of athletes and clusters. Each
athlete was considered as a cluster, with observed RTs
from both the `treatment' group (2022 World Championships) and
the `control' group (national competition) within the cluster. Cluster
sizes ranged from three to six, with a median size of four. Because gender
is known to influence RTs \citep{babicc2009reaction,
  lipps2011implications}, we prepared data for men and women
separately, resulting in 80 RTs from 17 athletes for each
gender. This setup forms the basis for the rank-based comparison for
clustered data with subunit-level grouping, which accounts for the
within-athlete dependence inherent in this data structure. The top
panel of Figure!~\ref{fig:RankScatterplots} shows the RTs
of those who competed both at national competitions and the 2022 World
Championships.

@@

Each athlete was treated as a single cluster, containing their reaction
times from different World Championships. The dataset for the 2019-2022
comparison contained 134 RTs from 34 male athletes and 124
RTs from 31 female athletes. The dataset for the 2022-2023
comparison contained 161 RTs from 45 male athletes and 182
RTs from 47 female athletes. While it is theoretically
possible that athletes improved their RTs between 2019 and
2022 or between 2022 and 2023, such improvements are highly unlikely for
elite sprinters, as they already operate near the limits of human
performance. Consequently, consistent improvements observed in 2022
would suggest systematic differences rather than natural variability.

@@

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subunit-level grouping. The test is designed based on a
within-cluster resampling principle. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the pseudo-sample.
The test statistic $S$ is the average of $W^*$ averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a $Z$ statistic which follows a standard normal distribution
asymptotically \citep[p.910]{datta2005rank}.

@@

The fitted parameters of the GG distribution int he
GAMLSS framework in Equations~\eqref{eq:gg}--\eqref{eq:sigma} are
summarized in Table~\ref{tab:ggfit}. Results obtained from both
excluding and including 2022 data are reported. The fixed-effect
parameters include $\beta_0$, $\gamma_0$, and $\nu$, corresponding to
the intercept of the log-location, log-scale, and shape of the GG
distribution, respectively. Random effects account for variability at
the venue level $\tau_v$ on the log-scale of the $\mu$ parameter
and at the heat level $\tau_h$ on the log-scale of the $\sigma$
parameter in the density in Equation~\eqref{eq:gg}. The variance
of the venue-level random effect is smaller than the heat-level random
effect variance, suggesting that heat-level variability in the scale
parameter is substantial, though on the dispersion parameter. When the
2022 data is included, all parameter estimates remain stable except
the standard deviation of the venue-level random-effect variance,
which increases from 0.043 to 0.058 . These results
highlight that RTs are influenced by both venue and heat-level
factors, and that the inclusion of 2022 introduces greater venue-level
variability, likely due to systematic differences in RTs that year.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{diagnosis.pdf}
  \caption{Diagnosis of the fitted GG distribution with
    random effects in model parameters: kernerl density of 1 million
    observations drawn from the fitted model overlaid with the
    histogram of the observed RTs (left); Q-Q plot of the
    normal z-score of the quantile residuals from the fitted model.}
  \label{fig:diagnosis}
\end{figure}

@@

\begin{table}
  \centering
  \caption{Probabilities of observing RTs less than threshold 0.08,
  0.09, and 0.10 seconds based on the two effect model.}
  \begin{tabular}{c c c c}
   \toprule
   Data Set & Threshold 0.08 & Threshold 0.09 & Threshold 0.10  \\
   \midrule
   Excluding 2022 & $5.31\cdot10^{-5}$ & $3.53\cdot10^{-4}$ &  $1.94\cdot10^{-3}$  \\
   Including 2022 & $6.84\cdot10^{-5}$ & $4.95\cdot10^{-4}$ & $2.76\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

@@

This study sought to answer two primary questions: were athletes who
competed at the 2022 World Track and Field Championships faster than
at other races, and is the 0.1-second RT barrier fair?
Regarding the first question, our analyses suggest that athletes were
indeed faster at the 2022 World Championships compared to comparable
meets. Ideally, a centralized database containing RTs from
all World Athletics-certified meets would have allowed for a more
comprehensive analysis. Currently, however, RTs are seldom
included in the results available on the World Athletics website. This
limitation constrained our ability to evaluate athletes like Devon
Allen, as data from his races leading up to the World Championships in
2022 were not accessible. A more robust dataset would enable deeper
insights by allowing comparisons across a larger set of races.

In addressing the second question, our analyses of the GAMLSS model
reveal that the 0.1-second RT barrier, while rare, may not
be as extraordinary  as traditionally assumed. For men,
Table~\ref{tab:Sim_time} shows that reaction  times below 0.1 seconds
occur with a probability of approximately one in 362 starts  when
including the 2022 data. Lowering the threshold to 0.08 seconds
drastically  reduces this likelihood, supporting the idea that the
current barrier could be adjusted to reflect more realistic
probabilities of false starts. A similar pattern is observed for
women, as detailed in the Supplementary Material, where reaction
times below 0.1 seconds are exceedingly rare for the 100-meter dash
and 100-meter  hurdles. However, the uniformity of the 0.1-second
barrier for both men and women  is questionable, given numerous
studies documenting gender differences in reaction  times
\citep[e.g.,][]{lipps2011implications, babicc2009reaction,
  panoutsakopoulos2020gender}. These studies suggest that the current
threshold may  unfairly penalize men, for whom sub-0.1-second reaction
times are more probable. \citet{brosnan2017effects} advocate for
gender-specific barriers, a position that  aligns with our findings
and highlights the importance of tailoring thresholds to  biological
distinctions.

>>>>

World Athletics (formerly IAAF) uses certified SIS to measure RTs, yet
variation in technology persists. Discussions at online forums such as
\url{www.LetsRun.com} questioned consistencies in the SIS as a
contributing factor to RT anormalies \citep{johnson2022data,
  johnson2022was}. Historically, ``loud gun'' systems caused signal delays
for athletes in outer lanes  due to the speed of sound, an issue
addressed with the introduction of ``silent gun'' systems in 2010,
which electronically synchronize sound delivery to all athletes
\citep{tonnessen2013reaction}. Despite these advances, variability
persists due to differences in sensor technologies, such as force
transducers and accelerometers, and inconsistencies in event detection
algorithms \citep{willwacher2013novel}. For example, simple
force-threshold systems may delay RT detection by up to 26~ms compared
to more sophisticated methods \citep{pain2007sprint}. These findings
emphasize the need for standardized certification protocols to reduce
discrepancies and ensure fairness in RT measurements, as recently
reviewed by \citet{milloz2021sprint}.

The 0.1-second disqualification threshold, introduced in the 1990s,
has been the subject of significant debate. Partly based on limited
data from Finnish national-level athletes~\cite{mero1990reaction}, the
threshold may not adequately represent the capabilities of elite
sprinters. Controlled experiments have shown that reaction times below
0.1 seconds are physiologically plausible \citep{pain2007sprint,
  komi2009iaaf}, while retrospective analyses of competition data
often advocate for raising the threshold~\citep{brosnan2017effects,
  lipps2011implications}. Stricter false-start rules, introduced to
minimize race disruptions, have also discouraged sprinters from
attempting faster starts, which may artificially inflate RTs recorded
in competition~\citep{haugen2013effect}. This divergence between
experimental findings and competition-based analyses illustrates the
complexity of defining a universally fair threshold. Addressing these
debates requires modern data collection and advanced methodologies to
ensure equity and consistency in elite
competition~\citep{milloz2021sprint}.

This paper addresses two primary objectives from a statistical
perspective, using modern methodologies to analyze historical
data. First, we investigate whether RTs at the 2022
World Championships were significantly different from other
competitions, focusing on athletes who competed in multiple
events. Using a matched-pairs design, we compare RTs across the 2022,
2019, and 2023 World Championships, as well as the 2022 national-level
competitions. This approach isolates the effect of the competition
year while controlling for individual performance. The clustered
nature of the data, with the goal of assessing differences across
events within athletes, was analyzed using a rank-based comparison
approach for clustered data \citep{datta2005rank}. Second, we evaluate
the appropriateness of the 0.1-second disqualification threshold by
modeling RTs from World Championships held from 1999 onward. A
generalized Gamma (GG) distribution with random effects for both venue and
heat was applied within the framework of the generalized additive
model for location, scale, and shape (GAMLSS)
\citep{rigby2005generalized, stasinopoulos2024generalized}.
This model enables estimation of the
probability of RTs falling below a threshold, providing a
statistical assessment of RT consistency and the validity of the
current threshold.

The rest of this paper is organized as follows. Section~\ref{sec:Data}
describes the data collection and datasets used for the two
objectives. Section~\ref{sec:rank} outlines the rank-based comparison
methods for objective~1, rank-based comparison for clustered data with
subunit-level grouping to analyze differences in RTs across different
competitions. Section~\ref{sec:gamlss} details the methods for
objective~2, the GG model with random effects to evaluate the appropriatenes
of the 0.1-second threshold. Section~\ref{sec:Results} presents the
results of both objectives and discusses the implications.
Finally, Section~\ref{sec:concludingremarks} highlights the
paper’s impact, limitations, and potential for future research.
All data and code for our analysis are provided in the supplementary
materials.

@@

The final dataset consisted of RTs from athletes who competed
in both national and international competitions. RTs from
preliminary heats, semifinals, and finals were included to ensure
sufficient data for analysis. Excluding preliminary heats would have
significantly reduced the number of athletes and clusters. Each
athlete was considered as a cluster, with observed RTs
from both the `treatment' group (2022 World Championships) and
the `control' group (national competition) within the cluster. Cluster
sizes ranged from three to six, with a median size of four. Because gender
is known to influence RTs \citep{babicc2009reaction,
  lipps2011implications}, we prepared data for men and women
separately, resulting in 80 RTs from 17 athletes for each
gender. This setup forms the basis for the rank-based comparison for
clustered data with subunit-level grouping, which accounts for the
within-athlete dependence inherent in this data structure. The top
panel of Figure~\ref{fig:RankScatterplots} shows the RTs
of those who competed both at national competitions and the 2022 World
Championships.

@@

Each athlete was treated as a single cluster, containing their RTs
from different World Championships. The dataset for the 2019-2022
comparison contained 134 RTs from 34 male athletes and 124
RTs from 31 female athletes. The dataset for the 2022-2023
comparison contained 161 RTs from 45 male athletes and 182
RTs from 47 female athletes. While it is theoretically
possible that athletes improved their RTs between 2019 and
2022 or between 2022 and 2023, such improvements are highly unlikely for
elite sprinters, as they already operate near the limits of human
performance. Consequently, consistent improvements observed in 2022
would suggest systematic differences rather than natural variability.

@@

\citet{datta2005rank} proposed an extension of the Wilcoxon rank-sum test to
clustered data with subunit-level grouping. The test is designed based on a
within-cluster resampling principle. Consider randomly picking one observation
from each cluster to form a pseudo-sample. Let $X_i^*$ be a random pick from the
$i$th cluster in the pseudo-sample and $\delta_i^*$ its group indicator. The
Wilcoxon rank-sum statistic for the pseudo-sample is
\[
W^* = \frac{1}{n + 1} + \sum_{i=1}^{n} \delta_{i}^{*} R_{i}^{*},
\]
where $R_{i}^{*}$ is the rank of $X_{i}^{*}$ in the pseudo-sample.
The test statistic $S$ is the average of $W^*$, averaged over all possible
pseudo-samples conditioning on the observed data and group indicators.
The mean and variance of $S$ under $H_0$ can be derived so that $S$ can be
standardized to form a $Z$ statistic which follows a standard normal distribution
asymptotically \citep[p.910]{datta2005rank}.

@@

The fitted parameters of the GG distribution in the
GAMLSS framework in Equations~\eqref{eq:gg}--\eqref{eq:sigma} are
summarized in Table~\ref{tab:ggfit}. Results obtained from both
excluding and including 2022 data are reported. The fixed-effect
parameters include $\beta_0$, $\gamma_0$, and $\nu$, corresponding to
the intercept of the log-location, log-scale, and shape of the GG
distribution, respectively. Random effects account for variability at
the venue level $\tau_v$ on the log-scale of the $\mu$ parameter
and at the heat level $\tau_h$ on the log-scale of the $\sigma$
parameter in the density in Equation~\eqref{eq:gg}. The variance
of the venue-level random effect is smaller than the heat-level random
effect variance, suggesting that heat-level variability in the scale
parameter is substantial, though on the dispersion parameter. When the
2022 data is included, all parameter estimates remain stable except
the standard deviation of the venue-level random-effect variance,
which increases from 0.043 to 0.058 . These results
highlight that RTs are influenced by both venue and heat-level
factors, and that the inclusion of 2022 introduces greater venue-level
variability, likely due to systematic differences in RTs that year.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{diagnosis.pdf}
  \caption{Diagnosis of the fitted GG distribution with
    random effects in model parameters: kernel density of 1 million
    observations drawn from the fitted model overlaid with the
    histogram of the observed RTs (left); Q-Q plot of the
    normal z-score of the quantile residuals from the fitted model.}
  \label{fig:diagnosis}
\end{figure}

@@

\begin{table}
  \centering
  \caption{Probabilities of observing RTs less than threshold 0.08,
  0.09, and 0.10 seconds based on the
    fitted GG GAMLSS model with both venue- and heat-level
random effects.}
  \begin{tabular}{c c c c}
   \toprule
   Data Set & Threshold 0.08 & Threshold 0.09 & Threshold 0.10  \\
   \midrule
   Excluding 2022 & $5.31\cdot10^{-5}$ & $3.53\cdot10^{-4}$ &  $1.94\cdot10^{-3}$  \\
   Including 2022 & $6.84\cdot10^{-5}$ & $4.95\cdot10^{-4}$ & $2.76\cdot10^{-3}$ \\
   \bottomrule
  \end{tabular}
  \label{tab:Sim_probability}
\end{table}

@@

This study sought to answer two primary questions: were athletes who
competed at the 2022 World Track and Field Championships faster than
at other races, and is the 0.1-second RT barrier fair?
Regarding the first question, our analyses suggest that athletes were
indeed faster at the 2022 World Championships compared to comparable
meets. 
\eds{are we testing faster (one-sided) or different (two-sided)?}
Ideally, a centralized database containing RTs from
all World Athletics-certified meets would have allowed for a more
comprehensive analysis. Currently, however, RTs are seldom
included in the results available on the World Athletics website. This
limitation constrained our ability to evaluate athletes like Devon
Allen, as data from his races leading up to the World Championships in
2022 were not accessible. A more robust dataset would enable deeper
insights by allowing comparisons across a larger set of races.

In addressing the second question, our analyses of the GAMLSS model
reveal that the 0.1-second RT barrier, while rare, may not
be as extraordinary  as traditionally assumed. For men,
Table~\ref{tab:Sim_time} shows that reaction  times below 0.1 seconds
occur with a probability of approximately one in 362 starts  when
including the 2022 data. Lowering the threshold to 0.08 seconds
drastically  reduces this likelihood, supporting the idea that the
current barrier could be adjusted to reflect more realistic
probabilities of false starts. A similar pattern is observed for
women, as detailed in the Supplementary Material, where reaction
times below 0.1 seconds are exceedingly rare for the 100-meter dash
and 100-meter  hurdles. However, the uniformity of the 0.1-second
barrier for both men and women  is questionable, given numerous
studies documenting gender differences in reaction  times
\citep[e.g.,][]{lipps2011implications, babicc2009reaction,
  panoutsakopoulos2020gender}. These studies suggest that the current
threshold may  unfairly penalize men, for whom sub-0.1-second reaction
times are more probable. \citet{brosnan2017effects} advocate for
gender-specific barriers, a position that  aligns with our findings
and highlights the importance of tailoring thresholds to  biological
distinctions. \eds{is this still true in the updated analyses?}
